---
title: "Validation: Stratified Sampler"
author: "rfaR Package Validation"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(rfaR)
```

## Purpose

Validate that `stratified_sampler()` correctly creates stratified bins and weights for all three supported distribution transformations: Extreme Value Type I (EV1), Normal, and Uniform. Tests cover bin boundary accuracy against the `stratified_example` dataset, weight accuracy, weight summation, output dimensions, and bin ordering.

## Background

Stratified sampling divides the probability space into bins to ensure adequate coverage of rare events. The choice of transformation determines how bins are distributed:

- **Uniform** — Equal probability width per bin. Simple but inefficient for rare event estimation.
- **Normal** — Uniform spacing in standard normal (z-score) space. Concentrates more bins near the center of the distribution.
- **EV1 (Gumbel)** — Spacing in Gumbel reduced variate space. Allocates progressively more bins to rare events, making it the recommended choice for flood frequency analysis.

---

## Test 1: Bin Spacing Visualization

The following plot illustrates how each transformation distributes bin boundaries along the z-variate axis. Tick marks show the z-lower boundary of each bin. EV1 stratification produces wider spacing at common events (left) and tighter spacing at rare events (right), concentrating sampling effort where it matters most for dam safety.

```{r bin-spacing-plot, fig.width=10, fig.height=5}
ev1 <- stratified_sampler(dist = "EV1")
normal <- stratified_sampler(dist = "Normal")
uniform <- stratified_sampler(dist = "Uniform")

# Shared x-axis range
xlim <- range(c(ev1$Zlower, ev1$Zupper,
                normal$Zlower, normal$Zupper,
                uniform$Zlower, uniform$Zupper))

par(mar = c(5, 6, 4, 2))
plot(NULL, xlim = xlim, ylim = c(0.5, 3.5), yaxt = "n",
     xlab = "Z-Variate (Standard Normal)", ylab = "",
     main = "Stratified Bin Spacing by Distribution Transformation")
axis(2, at = 1:3, labels = c("Uniform", "Normal", "EV1"), las = 1)
abline(h = 1:3, col = "gray90")

# Uniform bins
segments(uniform$Zlower, 0.8, uniform$Zlower, 1.2, col = "forestgreen", lwd = 2)
segments(uniform$Zupper[length(uniform$Zupper)], 0.8,
         uniform$Zupper[length(uniform$Zupper)], 1.2, col = "forestgreen", lwd = 2)

# Normal bins
segments(normal$Zlower, 1.8, normal$Zlower, 2.2, col = "steelblue", lwd = 2)
segments(normal$Zupper[length(normal$Zupper)], 1.8,
         normal$Zupper[length(normal$Zupper)], 2.2, col = "steelblue", lwd = 2)

# EV1 bins
segments(ev1$Zlower, 2.8, ev1$Zlower, 3.2, col = "coral", lwd = 2)
segments(ev1$Zupper[length(ev1$Zupper)], 2.8,
         ev1$Zupper[length(ev1$Zupper)], 3.2, col = "coral", lwd = 2)

legend("topleft", legend = c("Uniform", "Normal", "EV1"),
       col = c("forestgreen", "steelblue", "coral"), lwd = 2, bg = "white")
```

```{r weight-distribution-plot, fig.width=10, fig.height=5}
par(mfrow = c(1, 3), mar = c(5, 4, 3, 1))

barplot(uniform$Weights, names.arg = 1:uniform$Nbins, col = "forestgreen",
        main = "Uniform", xlab = "Bin", ylab = "Weight",
        ylim = c(0, max(c(uniform$Weights, normal$Weights, ev1$Weights)) * 1.1))

barplot(normal$Weights, names.arg = 1:normal$Nbins, col = "steelblue",
        main = "Normal", xlab = "Bin", ylab = "Weight",
        ylim = c(0, max(c(uniform$Weights, normal$Weights, ev1$Weights)) * 1.1))

barplot(ev1$Weights, names.arg = 1:ev1$Nbins, col = "coral",
        main = "EV1", xlab = "Bin", ylab = "Weight",
        ylim = c(0, max(c(uniform$Weights, normal$Weights, ev1$Weights)) * 1.1))
```

The weight distribution shows how each transformation allocates probability across bins. Uniform produces nearly equal weights. Normal concentrates weight in central bins. EV1 places the most weight in the first few bins (frequent events) with rapidly decreasing weight for rare event bins, reflecting the heavy-tailed nature of flood distributions.

---

## Test 2: Z-Lower Bin Boundaries Match Validation Data

Compare computed z-lower bin boundaries against the `stratified_example` dataset.

```{r test2-compute}
ev1_idx <- which(stratified_example$distribution == "ev1")
normal_idx <- which(stratified_example$distribution == "normal")
uniform_idx <- which(stratified_example$distribution == "uniform")

# Convert validation data to z-space
z_lower_valid <- numeric(nrow(stratified_example))
z_lower_valid[uniform_idx] <- qnorm(1 - stratified_example$lower[uniform_idx])
z_lower_valid[normal_idx] <- stratified_example$lower[normal_idx]
z_lower_valid[ev1_idx] <- qnorm(exp(-exp(-stratified_example$lower[ev1_idx])))

diff_ev1 <- ev1$Zlower - z_lower_valid[ev1_idx]
diff_normal <- normal$Zlower - z_lower_valid[normal_idx]
diff_uniform <- uniform$Zlower - z_lower_valid[uniform_idx]
```

```{r test2-table}
zlower_results <- data.frame(
  Distribution = c("EV1", "Normal", "Uniform"),
  Max_Abs_Difference = c(max(abs(diff_ev1)),
                          max(abs(diff_normal)),
                          max(abs(diff_uniform)))
)

knitr::kable(zlower_results, digits = 10,
             col.names = c("Distribution", "Max Absolute Difference"),
             caption = "Z-Lower Bin Boundary Differences vs. Validation Data")
```

### Acceptance Criterion

```{r test2-result}
tol_zlower <- 1e-6
pass_zlower <- all(c(max(abs(diff_ev1)),
                      max(abs(diff_normal)),
                      max(abs(diff_uniform))) < tol_zlower)
```

| Metric | Value |
|--------|-------|
| Tolerance | `r tol_zlower` |
| **Result** | **`r ifelse(pass_zlower, "PASS", "FAIL")`** |

---

## Test 3: Weights Match Validation Data

Compare computed bin weights against the `stratified_example` dataset.

```{r test3-compute}
wt_diff_ev1 <- ev1$Weights - stratified_example$weight[ev1_idx]
wt_diff_normal <- normal$Weights - stratified_example$weight[normal_idx]
wt_diff_uniform <- uniform$Weights - stratified_example$weight[uniform_idx]
```

```{r test3-table}
weight_results <- data.frame(
  Distribution = c("EV1", "Normal", "Uniform"),
  Max_Abs_Difference = c(max(abs(wt_diff_ev1)),
                          max(abs(wt_diff_normal)),
                          max(abs(wt_diff_uniform)))
)

knitr::kable(weight_results, digits = 10,
             col.names = c("Distribution", "Max Absolute Difference"),
             caption = "Weight Differences vs. Validation Data")
```

### Acceptance Criterion

```{r test3-result}
tol_wt <- 1e-6
pass_wt <- all(c(max(abs(wt_diff_ev1)),
                  max(abs(wt_diff_normal)),
                  max(abs(wt_diff_uniform))) < tol_wt)
```

| Metric | Value |
|--------|-------|
| Tolerance | `r tol_wt` |
| **Result** | **`r ifelse(pass_wt, "PASS", "FAIL")`** |

---

## Test 4: Weights Sum to 1

```{r test4-compute}
sum_ev1 <- sum(ev1$Weights)
sum_normal <- sum(normal$Weights)
sum_uniform <- sum(uniform$Weights)
```

```{r test4-table}
sum_results <- data.frame(
  Distribution = c("EV1", "Normal", "Uniform"),
  Weight_Sum = c(sum_ev1, sum_normal, sum_uniform),
  Difference_From_1 = c(abs(1 - sum_ev1), abs(1 - sum_normal), abs(1 - sum_uniform))
)

knitr::kable(sum_results, digits = 12,
             col.names = c("Distribution", "Sum of Weights", "|1 - Sum|"),
             caption = "Weight Summation Check")
```

### Acceptance Criterion

```{r test4-result}
tol_sum <- 1e-10
pass_sum <- all(c(abs(1 - sum_ev1), abs(1 - sum_normal), abs(1 - sum_uniform)) < tol_sum)
```

| Metric | Value |
|--------|-------|
| Tolerance | `r tol_sum` |
| **Result** | **`r ifelse(pass_sum, "PASS", "FAIL")`** |

---

## Test 5: Output Dimensions

Verify that `stratified_sampler()` returns the correct number of bins, events, and vector lengths.

```{r test5-compute}
test_custom <- stratified_sampler(Nbins = 10, Mevents = 100)

dim_results <- data.frame(
  Parameter = c("Nbins", "Mevents", "length(normOrd)", "length(Zlower)",
                "length(Zupper)", "length(Weights)"),
  Expected = c(10, 100, 1000, 10, 10, 10),
  Actual = c(test_custom$Nbins, test_custom$Mevents, length(test_custom$normOrd),
             length(test_custom$Zlower), length(test_custom$Zupper), length(test_custom$Weights))
)
```

```{r test5-table}
dim_results$Pass <- dim_results$Expected == dim_results$Actual

knitr::kable(dim_results,
             col.names = c("Parameter", "Expected", "Actual", "Pass"),
             caption = "Output Dimension Verification (Nbins=10, Mevents=100)")
```

### Acceptance Criterion

```{r test5-result}
pass_dims <- all(dim_results$Pass)
```

| Metric | Value |
|--------|-------|
| **Result** | **`r ifelse(pass_dims, "PASS", "FAIL")`** |

---

## Test 6: Bin Ordering (Zlower < Zupper)

Verify that all bin lower bounds are strictly less than upper bounds.

```{r test6-compute}
order_results <- data.frame(
  Distribution = c("EV1", "Normal", "Uniform"),
  All_Ordered = c(all(ev1$Zlower < ev1$Zupper),
                   all(normal$Zlower < normal$Zupper),
                   all(uniform$Zlower < uniform$Zupper))
)

knitr::kable(order_results,
             col.names = c("Distribution", "All Zlower < Zupper"),
             caption = "Bin Ordering Verification")
```

### Acceptance Criterion

```{r test6-result}
pass_order <- all(order_results$All_Ordered)
```

| Metric | Value |
|--------|-------|
| **Result** | **`r ifelse(pass_order, "PASS", "FAIL")`** |

---

## Summary

| Test | Description | Result |
|------|-------------|--------|
| 1 | Bin spacing visualization | (Visual) |
| 2 | Z-lower boundaries match validation data | **`r ifelse(pass_zlower, "PASS", "FAIL")`** |
| 3 | Weights match validation data | **`r ifelse(pass_wt, "PASS", "FAIL")`** |
| 4 | Weights sum to 1 | **`r ifelse(pass_sum, "PASS", "FAIL")`** |
| 5 | Output dimensions correct | **`r ifelse(pass_dims, "PASS", "FAIL")`** |
| 6 | Bin ordering (Zlower < Zupper) | **`r ifelse(pass_order, "PASS", "FAIL")`** |
