---
title: "Validation: Modified Puls Reservoir Routing"
author: "rfaR Package Validation"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(rfaR)
```

## Purpose

Validate that `mod_puls_routing()` produces results consistent with HEC-HMS Modified Puls routing for two benchmark datasets: John Martin Dam (JMD) and Cherry Creek Dam. Both benchmarks use identical reservoir models, inflow hydrographs, and initial conditions routed independently in HEC-HMS.

---

## Test 1: John Martin Dam

```{r jmd-compute}
jmd_routing <- mod_puls_routing(resmodel_df = jmd_resmodel,
                                inflow_df = jmd_inflowhydro,
                                initial_elev = jmd_init_elev,
                                full_results = TRUE)

jmd_diff_elev <- jmd_routing$elevation_ft - jmd_hms_results$elevation_ft
jmd_diff_outflow <- jmd_routing$outflow_cfs - jmd_hms_results$outflow_cfs

jmd_max_diff_elev <- max(abs(jmd_diff_elev))
jmd_max_diff_outflow <- max(abs(jmd_diff_outflow))
```

```{r jmd-table}
jmd_summary <- data.frame(
  Metric = c("Peak Elevation (ft)", "Peak Outflow (cfs)",
             "Max Abs Diff Elevation (ft)", "Max Abs Diff Outflow (cfs)"),
  rfaR = c(max(jmd_routing$elevation_ft), max(jmd_routing$outflow_cfs),
           jmd_max_diff_elev, jmd_max_diff_outflow),
  HMS = c(max(jmd_hms_results$elevation_ft), max(jmd_hms_results$outflow_cfs),
          NA, NA)
)

knitr::kable(jmd_summary, digits = 4,
             col.names = c("Metric", "rfaR", "HEC-HMS"),
             caption = "John Martin Dam Routing Comparison")
```

```{r jmd-plot, fig.width=10, fig.height=8}
par(mfrow = c(2, 1), mar = c(4, 4, 3, 1))

# Hydrograph comparison
plot(jmd_routing$time_hr, jmd_routing$inflow_cfs, type = "l", col = "gray70", lwd = 1,
     xlab = "Time (hr)", ylab = "Flow (cfs)",
     main = "John Martin Dam — Inflow and Outflow Comparison")
lines(jmd_routing$time_hr, jmd_routing$outflow_cfs, col = "steelblue", lwd = 2)
lines(jmd_hms_results$time_hr, jmd_hms_results$outflow_cfs, col = "coral", lwd = 2, lty = 2)
legend("topright", legend = c("Inflow", "rfaR Outflow", "HMS Outflow"),
       col = c("gray70", "steelblue", "coral"), lwd = c(1, 2, 2), lty = c(1, 1, 2))

# Difference time series
plot(jmd_routing$time_hr, jmd_diff_outflow, type = "l", col = "steelblue", lwd = 1,
     xlab = "Time (hr)", ylab = "Difference (cfs)",
     main = "John Martin Dam — Outflow Difference (rfaR - HMS)")
abline(h = 0, col = "gray50", lty = 2)
```

### Acceptance Criterion

Maximum absolute difference must be less than 0.1 ft for elevation and 1.0 cfs for outflow.

```{r jmd-result}
pass_jmd_elev <- jmd_max_diff_elev < 0.1
pass_jmd_outflow <- jmd_max_diff_outflow < 1.0
```

| Metric | Value | Tolerance |
|--------|-------|-----------|
| Max Abs Diff Elevation | `r round(jmd_max_diff_elev, 4)` ft | 0.1 ft |
| Max Abs Diff Outflow | `r round(jmd_max_diff_outflow, 4)` cfs | 1.0 cfs |
| **Result (Elevation)** | **`r ifelse(pass_jmd_elev, "PASS", "FAIL")`** | |
| **Result (Outflow)** | **`r ifelse(pass_jmd_outflow, "PASS", "FAIL")`** | |

---

## Test 2: Cherry Creek Dam

```{r cc-compute}
cc_routing <- mod_puls_routing(resmodel_df = cc_resmodel,
                               inflow_df = cc_inflowhydro,
                               initial_elev = cc_init_elev,
                               full_results = TRUE)

cc_diff_elev <- cc_routing$elevation_ft - cc_hms_results$elevation_ft
cc_diff_outflow <- cc_routing$outflow_cfs - cc_hms_results$outflow_cfs

cc_max_diff_elev <- max(abs(cc_diff_elev))
cc_max_diff_outflow <- max(abs(cc_diff_outflow))
```

```{r cc-table}
cc_summary <- data.frame(
  Metric = c("Peak Elevation (ft)", "Peak Outflow (cfs)",
             "Max Abs Diff Elevation (ft)", "Max Abs Diff Outflow (cfs)"),
  rfaR = c(max(cc_routing$elevation_ft), max(cc_routing$outflow_cfs),
           cc_max_diff_elev, cc_max_diff_outflow),
  HMS = c(max(cc_hms_results$elevation_ft), max(cc_hms_results$outflow_cfs),
          NA, NA)
)

knitr::kable(cc_summary, digits = 4,
             col.names = c("Metric", "rfaR", "HEC-HMS"),
             caption = "Cherry Creek Dam Routing Comparison")
```

```{r cc-plot, fig.width=10, fig.height=8}
par(mfrow = c(2, 1), mar = c(4, 4, 3, 1))

# Hydrograph comparison
plot(cc_routing$time_hr, cc_routing$inflow_cfs, type = "l", col = "gray70", lwd = 1,
     xlab = "Time (hr)", ylab = "Flow (cfs)",
     main = "Cherry Creek Dam — Inflow and Outflow Comparison")
lines(cc_routing$time_hr, cc_routing$outflow_cfs, col = "steelblue", lwd = 2)
lines(cc_hms_results$time_hr, cc_hms_results$outflow_cfs, col = "coral", lwd = 2, lty = 2)
legend("topright", legend = c("Inflow", "rfaR Outflow", "HMS Outflow"),
       col = c("gray70", "steelblue", "coral"), lwd = c(1, 2, 2), lty = c(1, 1, 2))

# Difference time series
plot(cc_routing$time_hr, cc_diff_outflow, type = "l", col = "steelblue", lwd = 1,
     xlab = "Time (hr)", ylab = "Difference (cfs)",
     main = "Cherry Creek Dam — Outflow Difference (rfaR - HMS)")
abline(h = 0, col = "gray50", lty = 2)
```

### Acceptance Criterion

Maximum absolute difference must be less than 0.1 ft for elevation and 1.0 cfs for outflow.

```{r cc-result}
pass_cc_elev <- cc_max_diff_elev < 0.1
pass_cc_outflow <- cc_max_diff_outflow < 1.0
```

| Metric | Value | Tolerance |
|--------|-------|-----------|
| Max Abs Diff Elevation | `r round(cc_max_diff_elev, 4)` ft | 0.1 ft |
| Max Abs Diff Outflow | `r round(cc_max_diff_outflow, 4)` cfs | 1.0 cfs |
| **Result (Elevation)** | **`r ifelse(pass_cc_elev, "PASS", "FAIL")`** | |
| **Result (Outflow)** | **`r ifelse(pass_cc_outflow, "PASS", "FAIL")`** | |

---

## Summary

| Test | Dam | Variable | Result |
|------|-----|----------|--------|
| 1a | John Martin | Elevation | **`r ifelse(pass_jmd_elev, "PASS", "FAIL")`** |
| 1b | John Martin | Outflow | **`r ifelse(pass_jmd_outflow, "PASS", "FAIL")`** |
| 2a | Cherry Creek | Elevation | **`r ifelse(pass_cc_elev, "PASS", "FAIL")`** |
| 2b | Cherry Creek | Outflow | **`r ifelse(pass_cc_outflow, "PASS", "FAIL")`** |
