---
title: "Setting Up Your rfaR Project"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting Up Your rfaR Project}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setting Up Your rfaR Project

We recommend organizing your analysis as an RStudio project. This keeps
file paths relative and reproducible across machines.

### Step 1 — Create a new RStudio Project

In RStudio: **File → New Project → New Directory → New Project**

Name it something like `my_dam_rfa/`. This creates a `.Rproj` file and
sets the working directory automatically on open.

### Step 2 — Recommended directory structure
```
my_dam_rfa/
├── my_dam_rfa.Rproj
├── data/
│   ├── resmodel.csv
│   ├── hydrographs/
│   │   ├── event_1.csv
│   │   └── event_2.csv
│   └── stage_record.csv
├── R/
│   └── my_dam_simulate.R
└── output/
```

### Step 3 — Import your data

`rfaR` expects data in specific formats. The built-in datasets show
the required structure:
```r
# Reservoir model: elevation, storage, outflow
head(jmd_resmodel)

# Hydrograph: time and flow columns
head(jmd_hydro_apr1999)

# Stage record: date and stage columns  
head(jmd_wy1980_stage)
```

Read your own data using `read.csv()` with relative paths:
```r
resmodel  <- read.csv("data/resmodel.csv")
stage_rec <- read.csv("data/stage_record.csv")
```

### Step 4 — Verify column structure

Before running `rfa_simulate()`, verify your imported data matches
the expected format:
```r
# Check against built-in example
str(jmd_resmodel)
str(resmodel)  # should match
```

```{r setup}
library(rfaR)
```
