---
title: "Conceptual Computation of an rfaR Realization"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Conceptual Computation of an rfaR Realization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.align = 'center',
  comment = "#>")
```

## Executive Summary

RMC-RFA produces a reservoir stage-frequency curve with uncertainty bounds by utilizing a deterministic flood routing model while treating the inflow volume, flood hydrograph shape, seasonal occurrence, and antecedent reservoir stage as uncertain variables rather than fixed values. To quantify both the natural variability and epistemic uncertainty in reservoir stage-frequency estimates, RMC-RFA employs a two-looped, nested Monte Carlo methodology.

The inner loop — defined here as a realization — simulates natural variability by routing 10,000 individual flood events through the reservoir, each with randomly sampled starting stage, hydrograph shape, and stratified-sampled inflow volume drawn from the volume-frequency distribution. Together these routed events produce a single stage-frequency curve estimate. The outer loop propagates epistemic uncertainty by repeating this process across many realizations, each conditioned on a different set of volume-frequency distribution parameters drawn from the posterior.

This document walks through a single realization in detail. Understanding one realization is the key to understanding the full simulation: the outer loop simply repeats this process with a resampled parameter set each time (in `rfaR`, this is the *next* parameter set from the input). The walkthrough covers the stratified sampling procedure, random sampling of starting stage and hydrograph shape, and calculation of stage exceedance probabilities — steps that are common to all simulation types. The full uncertainty simulation conducts 10,000 realizations, each comprising 10,000 flood events (100 million routing operations total), while the expected value simulation conducts 10,000 realizations, each using a one of the 10,000 BestFit parameter sets.

***Note:*** Users with coding experience may find the loops in this script "inefficient". This purpose of this step-by-step guide is to provide a replicable guide to the realization process.

```{r setup, echo = FALSE, warning = FALSE, message = FALSE}
library(rfaR)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggsci)
```

## `rfaR` Analysis Summary

A full uncertainty run in `rfa_simulate()` is comprised of 10,000 realizations. Each realization is compromised of 10,000 reservoir routing computations (simulations) using the following:

-   Stratified samples (10,000 samples comprised of 20 bins with 500 events each) from an inflow volume-frequency curve (VFC). The VFC is estimated using one parameter set

-   10,000 Seasonality samples represented by the month

-   10,000 Starting stage samples, dependent on the sampled month

-   Hydrograph Shape, scaled to sampled inflow volume

-   Modified-Puls routings to obtain peak stages and/or peak discharge

These module emulate the workflow and methodology from RMC-RFA. The subsections below will step through an rfaR realization. These sections represent modules and functions contained in `rfa_simulate()`.

## Flow-Frequency Stratified Sampling

### Stratified Bins Definition & Set-Up

Stratified sampling is used instead of simple random sampling because reliable estimates of exceedance probabilities far into the tail of the distribution — events with AEPs of 1/10,000 or rarer — are required. Simple random sampling would require an impractically large number of events to adequately populate the tail. Instead, the frequency axis is divided into `Nbins` equal-probability bins, and `Mevents` are sampled uniformly within each bin. This ensures the tail is well-represented without artificially inflating the total number of routing operations.

```{r}
ords <- stratified_sampler(Nbins = 20,
                               Mevents = 500,
                               dist = "ev1")
```

The bin weights (`ords$Weights`) reflect the true probability mass that each bin represents. These weights are applied later to reconstruct the true exceedance probability from the stratified sample, correcting for the fact that rare-event bins were intentionally oversampled relative to their actual frequency of occurrence.

### Construct a Probability Matrix from the Stratified Bins

The z-variate matrix translates each stratified sample into a standard normal deviate. Each column corresponds to one bin; each row is one of the `Mevents` random draws within that bin. Because EV1 (Gumbel) space is used for stratification, the bin boundaries (`Zlower`, `Zupper`) are in defined in Gumbel-reduced-variate space, and events are drawn uniformly within each bin. The resulting `z_matrix` is an `Mevents` × `Nbins` matrix of z-variates, with each column densely and uniformly populated across its corresponding probability/z-variate range.

```{r}
z_matrix <- matrix(ncol = ords$Nbins, nrow = ords$Mevents)

# Using EV1
for (i in 1:ords$Nbins){
  # Lower Bin Boundary (prior bin's upper boundary)
  bin_lower <- ords$Zlower[i]
  
  # Upper Bin Boundary
  bin_upper <- ords$Zupper[i]

  # Vector of random values
  z_matrix[,i] <- (bin_lower + (runif(500, min = 0, max = 1))*(bin_upper - bin_lower))
}
```

### Construct a Matrix of Corresponding Flow Quantiles

The Z-variates are converted to inflow volumes using the LP3 (Log-Pearson Type III) quantile function. The conversion routes through normal probability space: Z-variates → normal CDF (`pnorm`) → LP3 quantile (`qp3`). The result is `Q_matrix`, which has the same dimensions as `z_matrix` and each cell holds an inflow volume. Each inflow volume corresponds to a sample from the inflow volume-frequency distribution.

The parameter set used here (`meanlog`, `sdlog`, `skewlog`) represents a single draw from the distribution of LP3 parameters — one posterior parameter set for this realization.

```{r}
meanlog <- 3.550399
sdlog <- 0.3717982
skewlog <- 0.7555138

Q_matrix <- 10^qp3(pnorm(z_matrix), meanlog, sdlog, skewlog)
```

```{r stratified-bin-samples-plot, fig.width = 7, fig.height = 5, echo = FALSE, warning = FALSE}
# Compute the 21 boundary AEP values
aep_bounds <- 1 - pnorm(unique(c(ords$Zlower, ords$Zupper)))

# Format nicely in scientific notation
aep_labels <- formatC(aep_bounds, format = "E", digits = 1)

aep_major <- c(9.9e-01, 9.0e-01, 5.0e-01, 1.0e-01, 1.0e-02, 
               1.0e-03, 1.0e-04, 1.0e-05, 1.0e-06, 1.0e-07, 1.0e-08)

as.data.frame(Q_matrix) |>
  setNames(1:20) |>
  pivot_longer(everything(), names_to = "bin", values_to = "Q") |>
  mutate(bin = factor(as.integer(bin))) |>
  ggplot(aes(x = bin, y = Q)) +
  geom_jitter(aes(color = bin), size = 0.5, alpha = 0.7) +
  scale_color_viridis_d(option = "turbo") +
  scale_x_discrete(breaks = 1:20, labels = aep_labels[1:20]) +
  scale_y_log10(breaks = scales::log_breaks(n = 5, base = 10),
                minor_breaks = scales::minor_breaks_log(detail = 1)) +
  theme_rfar_conceptual()+
  #theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7)) +
  labs(x = "Bin Midpoint AEP", 
       y = "Inflow Volume (cfs)",
       color = NULL,
       title = "Inflow Volume-Frequency Sampling",
       subtitle = "LP3: (3.5504, 0.3718, 0.7555)")

```

## Define the Number of Simulations (Routings)

The total number of inflow events in this realization is the product of the number of bins and the number of events per bin — in this case, 20 × 500 = 10,000. This is the number of reservoir routing simulations that will be executed in the realization. The default is 10,00 simulations (equal to the number of parameter sets). Currently, there is no option to change the number of simulations or realizations.

```{r}
Nsims <- nrow(Q_matrix) * ncol(Q_matrix)
```

## Sample Seasonality

The month of flood occurrence is sampled randomly using historical relative frequencies from the site's seasonal record. This captures the seasonal pattern of large inflow or high stage events. The sampled month is used in the next step to sample an observed starting reservoir stage that is climatologically consistent with the inflow timing, preserving the correlation between stage elevation and seasonality.

The sample is a pre-allocated vector/sequence of months that is `Nsims` long.

```{r}
InitMonths <- sample(1:12, size = Nsims, replace = TRUE, prob = jmd_seasonality$relative_frequency)

UniqMonths <- sort(unique(InitMonths))

```

```{r init_month-seq-plot, fig.width = 7, fig.height = 5, echo = FALSE, warning = FALSE}
InitMonths_df <- data.frame(Sim_n = seq(1,length(InitMonths),1),
                            InitMonths = InitMonths)
ggplot(InitMonths_df) + 
  geom_bar(aes(factor(as.integer(InitMonths)), 
               fill = factor(as.integer(InitMonths))),
           color = "grey80") + 
  ggsci::scale_fill_aaas() +
  scale_x_discrete() + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) + 
  labs(x = "Sampled Month", y = "Count within Sample", fill = NULL,
       title = "Seasonality Sampling",
       subtitle = "Total Count of Monthly Samples") +
  theme_rfar_conceptual()

```

## Sample Starting Stage

The antecedent reservoir stage for each event is sampled conditionally on the sampled month of occurrence. Observed stages are sampled from the month they occurred, using the sampled flood month in the step prior. This ensures that starting conditions are seasonally realistic — an event sampled in April will begin from a stage elevation sampled from April in the observed stage time series. The starting stages are therefore sampled proportionately to their historic frequency of occurrence.

The sample is a pre-allocated vector/sequence of starting stages that is `Nsims` long.

```{r}
InitStages <- numeric(Nsims)

stage_ts <- jmd_wy1980_stage
stage_ts$months <- lubridate::month(lubridate::mdy(stage_ts$date))

for (i in 1:length(UniqMonths)) {
  sampleID <- which(InitMonths == UniqMonths[i])
  InitStages[sampleID] <- sample(stage_ts$stage[stage_ts$months %in% UniqMonths[i]],
                                 size = sum(InitMonths == UniqMonths[i]), replace = TRUE)
}

```

```{r init_stage-seq-plot, fig.width = 7, fig.height = 5, echo = FALSE, warning = FALSE}
InitStages_df <- data.frame(Sim_n = seq(1,length(InitMonths),1),
                            InitMonths = InitMonths,
                            InitStages = InitStages)

ggplot(InitStages_df) +
  geom_jitter(aes(x = InitMonths, InitStages, color = factor(as.integer(InitMonths))), size = 1, alpha = 0.7) +
  ggsci::scale_color_aaas() +
  scale_y_continuous(breaks = seq(3790,3900,10),
                     labels = scales::label_comma()) + 
  scale_x_continuous(breaks = seq(4,9,1)) +
  coord_cartesian(ylim = c(3790,3870)) + 
  labs(x = "Sampled Month", y = "Sampled Starting Stage", 
       color = NULL,
       title = "Samples of Starting Stage Elevations",
       subtitle = "Samples are taken from observed stage timeseries") +
  theme_rfar_conceptual()

```

## Sample Hydrograph Shape

The shape of the flood hydrograph is sampled randomly from a selection of historical and synthetic hydrographs. Each hydrograph in the library is assigned a weight (default weights are uniform); the sampled shape is then scaled to match the flood volume drawn from the LP3 distribution. Note that the defined critical duration is 2-days (`critical_duration`) and the routing window has been set to 10-days (`routing_days`).

```{r}
hydrographs <- hydrograph_setup(jmd_hydro_apr1999,
                                jmd_hydro_jun1921,
                                jmd_hydro_jun1965,
                                jmd_hydro_jun1965_15min,
                                jmd_hydro_may1955,
                                jmd_hydro_pmf,
                                jmd_hydro_sdf,
                                critical_duration = 2,
                                routing_days = 10)

hydroSamps <- sample(1:length(hydrographs), size = Nsims, replace = TRUE,
                     prob = attr(hydrographs, "probs"))
```

## Conduct Reservoir Routing Simulations

With all random variables sampled, the routing loop executes the Modified Puls method for each of the `Nsims` events. For each event, three inputs are combined: (1) the inflow hydrograph, scaled from the sampled shape to match the LP3-derived inflow volume; (2) the initial reservoir elevation drawn for that event; and (3) the reservoir's storage-outflow relationship (`resmodel_df`). The Modified Puls method numerically solves the storage equation at each timestep, tracking the reservoir elevation through the duration of the flood. Only the peak stage reached during the event is retained in this example, but peak discharge is also available.

The loop is structured to preserve the bin (column) alignment of `Q_matrix`: event `(i, j)` in `peakStage` corresponds to bin `j` and event `i`. Maintaining this alignment is essential for correctly applying the stratified bin weights during the stage-frequency estimation.

```{r}
peakStage <- matrix(NA, nrow = nrow(Q_matrix), ncol = ncol(Q_matrix))

realiz <- 0

for (i in 1:nrow(Q_matrix)) {
  for (j in 1:ncol(Q_matrix)) {
    # Keeps the indices lined up - necessary for the weights later
    realiz <- (i - 1) * ncol(Q_matrix) + j

    # Hydrograph shape
    hydrograph_shape <- hydrographs[[hydroSamps[realiz]]][,2:3]

    # Hydrograph observed volume
    obs_hydrograph_vol <- attr(hydrographs[[hydroSamps[realiz]]],"obs_vol")

    # Scale hydrograph to sample volume
    scaled_hydrograph <- scale_hydrograph(hydrograph_shape,
                                          obs_hydrograph_vol,
                                          Q_matrix[i,j])

    # Route scaled hydrograph
    tmpResults <- mod_puls_routing(resmodel_df = jmd_resmodel,
                                   inflow_df = scaled_hydrograph,
                                   initial_elev = InitStages[realiz],
                                   full_results = FALSE)

    # Record results
    peakStage[i, j] <- tmpResults[1]
  }
}
```

## Realizaton Results

### Determine Min and Max Stages

The range of routed peak stages defines the domain over which the realization stage-frequency curve will be estimated. All subsequent exceedance probability calculations (within the realization) are bounded by these values.

```{r}
min_stage <- min(peakStage)
max_stage <- max(peakStage)
```

### Define Exceedance Stages

A vector of stage thresholds is constructed by stepping evenly & incrementally from the minimum to the maximum peak stage across the number of events in each bin (`Mevents`). Each threshold in `stage_vect` represents a candidate stage level at which exceedance probability will be computed, forming the vertical axis of the stage-frequency curve.

For consistency with code chunks, the candidate stages in `stage_vect` can be referred to as "exceedance stages". Note that the vector of exceedance stages can be any length (ex. `n_exceedance_stages` could be 1,000).

```{r}
n_exceedance_stages <- ords$Mevents
stage_vect <- rep(NA, n_exceedance_stages)

for (b in 1:(n_exceedance_stages)){
  # Min Stage
  if(b < 2){
    stage <- min_stage
    
  # Incrementally calculate the next exceedance stage  
  }else{
    stage <- stage_vect[b-1] + (max_stage - min_stage)/(n_exceedance_stages)
  }
  stage_vect[b] <- stage
}
```

### Count Exceedances & Relative Frequencies

For each bin (column) of `peakStage` and each exceedance stage (`exceedance_stage`), the number of events with peak stages greater than the exceedance stage is totaled and divided by the number of events. This is the probability of exceeding a stage, given that the inflow volume was drawn from that inflow volume-frequency bin. The result is stored in `stage_exceedance_matrix`, with rows corresponding to exceedance stages in `stage_vect` and columns corresponding to stratified bins.

The accompanying graphic illustrates this process. In each bin, the number of stages exceeding `stage_vect[1]` is summed by `stage_exceed_count`. This is then divided by the number of routed events in the bin (`mstages`) to represent a probability of exceedance (`stage_exceed_prob`). This processes is repeated for each exceedance stage in `stage_vect`, within each bin.

```{r}
stage_exceedance_matrix <- matrix(nrow = length(stage_vect), ncol = ncol(Q_matrix))

for(m in 1:ncol(peakStage)){
  # Get Bin of peak stages, each represented by m event
  mstages <- peakStage[,m]

  for(n in 1:length(stage_vect)){
    # Get Stage
    exceedance_stage <- stage_vect[n]

    # How many exceedances in Bin
    stage_exceed_count <- sum(mstages > exceedance_stage)

    # Exceedance Prob in Bin
    stage_exceed_prob <- stage_exceed_count/length(mstages)

    # Save to exceedance matrix 
    stage_exceedance_matrix[n,m] <- stage_exceed_prob
  }
}
```

```{r peak_stage_binned-plot, fig.width = 7, fig.height = 5, echo = FALSE, warning = FALSE}
as.data.frame(peakStage) |>
  setNames(1:20) |>
  pivot_longer(everything(), names_to = "bin", values_to = "peak_stage") |>
  mutate(bin = factor(as.integer(bin))) |>
  ggplot(aes(x = bin, y = peak_stage)) +
  geom_jitter(aes(color = bin), size = 0.9, alpha = 0.7) +
  scale_color_viridis_d(option = "turbo") +
  scale_x_discrete(breaks = 1:20, labels = aep_labels[1:20]) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10),
                     labels = scales::label_comma()) +
  labs(x = "Bin Midpoint AEP", 
       y = "Peak Stage (ft-NAVD88)",
       color = NULL,
       title = "Peak Stage Results by Stratified Bin",
       subtitle = "Routing results from 10,000 events") + 
  theme_rfar_conceptual() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7)) 

```

## Estimate Exceedance Probabilities using Stratified Bin Weights

The stage exceedance probabilities from each bin are combined using the stratified bin weights. The weighted sum of bin-level exceedance probabilities estimates the annual exceedance probability (AEP) at a given exceedance stage. The weighted sum is to correct for the intentional oversampling of rare/less-frequent inflow events. For more on this see the [Law of Total Probability](https://en.wikipedia.org/wiki/Law_of_total_probability).

The result is `stage_result` which is the stages from `stage_vect` and their corresponding AEP estimates. This is the realization stage-frequency curve. This curve can be interpolated using stages or AEPs.

```{r}
stage_aep_vect <- rep(NA,length(stage_vect))

for(m in 1:nrow(stage_exceedance_matrix)){
  # Grab row of exceedances
  stage_exceedance_probs <- stage_exceedance_matrix[m,]

  # Sum the dot product of exceedances and weights
  stage_aep <- sum(stage_exceedance_probs*ords$Weights)

  #save to vector
  stage_aep_vect[m] <- stage_aep
}

stage_result <- tibble(
  AEP = stage_aep_vect,
  Z_var = qnorm(1-stage_aep_vect),
  Gumb = -log(-log(1 - AEP)),
  Stage = stage_vect)

```

```{r bin_weighed_prob-plot, fig.width = 7, fig.height = 5, echo = FALSE, warning = FALSE}
# Reshape exceedance matrix to long
stage_exc_long <- as.data.frame(stage_exceedance_matrix) |>
  setNames(1:20) |>
  mutate(Stage = stage_vect) |>
  pivot_longer(-Stage, names_to = "bin", values_to = "AEP") |>
  mutate(bin = as.integer(bin))

near_1 <- qnorm(1 - 0.999999999999)
near_0 <- qnorm(1-1e-16)

ggplot() +
  geom_line(data = stage_exc_long,
            aes(x = qnorm(1 - AEP), y = Stage, color = factor(bin)),
            linewidth = 0.75, 
            alpha = 0.6) +
  geom_line(data = stage_result,
            aes(x = Z_var, y = Stage),
            linewidth = 1.25, color = "black") +
  scale_x_continuous(breaks = qnorm(1 - aep_major),
                     labels = formatC(aep_major, format = "E", digits = 0)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10),
                     labels = scales::label_comma()) +
  scale_color_viridis_d(option = "turbo") +
  coord_cartesian(xlim = c(near_1, near_0)) +
  theme_rfar_conceptual()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(x = "AEP", y = "Peak Stage (ft-NAVD88)",
       title = "Stage-Frequency Curve",
       subtitle = "Colored lines = per-bin exceedance curves; black = weighted results (sum of weighted bin exceedance probabilities)")


```
