---
title: "Starting Stage Sampling Validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Starting Stage Sampling Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rfaR)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Purpose

Validate the starting pool elevation sampling module for Jay McGraw Dam (JMD). This involves two tests:

1. **Stage Duration Curve Verification** — Confirm that quantiles computed directly from the observed stage time series match the `jmd_stage_duration` dataset included in the package.

2. **Sampled Stage Validation** — Confirm that stages sampled via the seasonality-weighted sampling process reproduce the observed stage duration curves within acceptable tolerance.

## Input Data

The observed stage record (`jmd_wy1980_stage`) contains daily pool elevation readings. The `jmd_stage_duration` dataset contains pre-computed monthly stage duration curves at specified exceedance probabilities.

```{r input-data}
stage_ts <- jmd_wy1980_stage
stage_ts$months <- lubridate::month(lubridate::mdy(jmd_wy1980_stage$date))
stage_ts <- stage_ts[!is.na(stage_ts$months), ]

probs <- 1 - jmd_stage_duration$Probability
```

---

## Test 1: Observed Quantiles vs. Package Dataset

Compute stage quantiles directly from the observed time series for each month and compare against `jmd_stage_duration`.

```{r test1-compute}
obs_quantiles <- sapply(1:12, function(m) {
  unname(quantile(stage_ts$stage_ft[stage_ts$months == m],
                  probs, na.rm = TRUE))
})

sd_matrix <- as.matrix(jmd_stage_duration[, -1])
diff_matrix <- obs_quantiles - sd_matrix
max_diff_obs <- max(abs(diff_matrix), na.rm = TRUE)
```

```{r test1-table}
test1_summary <- data.frame(
  Month = month.abb,
  Max_Abs_Difference = apply(abs(diff_matrix), 2, max, na.rm = TRUE))

knitr::kable(test1_summary, digits = 4,
             col.names = c("Month", "Max Absolute Difference (ft)"),
             caption = "Maximum Difference Between Computed and Package Stage Duration Curves")
```

### Acceptance Criterion

Maximum absolute difference must be less than 0.01 ft, consistent with `expect_lt(max_diff, 0.01)`.

```{r test1-result}
pass_test1 <- max_diff_obs < 0.01
```

| Metric | Value |
|--------|-------|
| Maximum Absolute Difference | `r round(max_diff_obs, 6)` ft |
| Tolerance | 0.01 ft |
| **Result** | **`r ifelse(pass_test1, "PASS", "FAIL")`** |

---

## Test 2: Sampled Stage Quantiles vs. Observed Quantiles

Sample 10,000 months using the JMD seasonality distribution, then for each sampled month, sample a starting stage from the observed record for that month. Compare the resulting stage duration curves against the observed curves.

```{r test2-sampling}
set.seed(42)

seasonality_prob <- jmd_seasonality$relative_frequency
Nsims <- 10000

InitMonths <- sample(1:12, size = Nsims, replace = TRUE, prob = seasonality_prob)
InitStages <- numeric(Nsims)

UniqMonths <- sort(unique(InitMonths))
for (i in 1:length(UniqMonths)) {
  sampleID <- which(InitMonths == UniqMonths[i])
  InitStages[sampleID] <- sample(stage_ts$stage_ft[stage_ts$months %in% UniqMonths[i]],
                                 size = sum(InitMonths == UniqMonths[i]), replace = TRUE)
}

sample_stages <- data.frame(months = InitMonths, stage_ft = InitStages)
```

```{r test2-compute}
sample_quantiles <- sapply(1:12, function(m) {
  unname(quantile(sample_stages$stage_ft[sample_stages$months == m],
                  probs, na.rm = TRUE))
})

sampled_months <- which(!is.na(sample_quantiles[1, ]))
unsampled_months <- which(is.na(sample_quantiles[1, ]))

pct_diff_matrix <- abs((sample_quantiles[, sampled_months] -
                         obs_quantiles[, sampled_months]) /
                        obs_quantiles[, sampled_months]) * 100

max_pct_diff <- max(pct_diff_matrix, na.rm = TRUE)
```

```{r test2-table}
test2_summary <- data.frame(
  Month = month.abb[sampled_months],
  Max_Pct_Difference = apply(pct_diff_matrix, 2, max, na.rm = TRUE)
)

knitr::kable(test2_summary, digits = 4,
             col.names = c("Month", "Max Percent Difference (%)"),
             caption = "Maximum Percent Difference Between Sampled and Observed Stage Quantiles")
```

**Months not sampled** (zero seasonality probability): `r paste(month.abb[unsampled_months], collapse = ", ")`. These months correctly produce no samples.

```{r test2-plot, fig.width=10, fig.height=6}
# Build observed quantiles long
obs_long <- as.data.frame(obs_quantiles) |>
  setNames(month.abb) |>
  mutate(prob = probs) |>
  pivot_longer(cols = -prob,
               names_to  = "Month",
               values_to = "Stage") |>
  mutate(Type = "Observed")

# Build sampled quantiles long
samp_long <- as.data.frame(sample_quantiles) |>
  setNames(month.abb) |>
  mutate(prob = probs) |>
  pivot_longer(cols = -prob,
               names_to  = "Month",
               values_to = "Stage") |>
  mutate(Type = "Sampled")

# Combine and factor
sampled_months <- (unique(samp_long$Month[!is.na(samp_long$Stage)]))
months_i_want <- rep(NA,length.out = length(sampled_months))

for (i in 1:length(sampled_months)){
  months_i_want[i] <- month.abb[sampled_months[i]]
}

quantile_long <- bind_rows(obs_long, samp_long) |>
  filter(!is.na(Stage)) |>
  filter(Month %in% unique(samp_long$Month[!is.na(samp_long$Stage)])) |> 
  mutate(Month = factor(Month, levels = month.abb),
         Type  = factor(Type, levels = c("Observed", "Sampled")))

ggplot(quantile_long, aes(x = prob, y = Stage,
                          color = Type, linetype = Type)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~Month, nrow = 3) +
  scale_x_reverse() +
  scale_color_manual(values = c("Observed" = "#3B4992FF",
                                "Sampled"  = "#EE0000FF")) +
  scale_linetype_manual(values = c("Observed" = "solid",
                                   "Sampled"  = "dashed")) +
  labs(x = "Exceedance Probability",
       y = "Stage (ft)",
       color = NULL,
       linetype = NULL) +
  theme_bw() +
  theme(legend.position = "top",
        panel.grid.minor = element_blank(),
        strip.background = element_blank(),
        strip.text = element_text(face = "bold"))
```

### Acceptance Criterion

Maximum percent difference between sampled and observed stage quantiles must be less than 5%, consistent with `expect_lt(pct_diff, 5)`.

```{r test2-result}
pass_test2 <- max_pct_diff < 5
```

| Metric | Value |
|--------|-------|
| Maximum Percent Difference | `r round(max_pct_diff, 4)`% |
| Tolerance | 5% |
| **Result** | **`r ifelse(pass_test2, "PASS", "FAIL")`** |

---

## Summary

| Test | Description | Result |
|------|-------------|--------|
| 1 | Observed quantiles match `jmd_stage_duration` | **`r ifelse(pass_test1, "PASS", "FAIL")`** |
| 2 | Sampled stage quantiles within 5% of observed | **`r ifelse(pass_test2, "PASS", "FAIL")`** |
