---
title: "V2 - BestFit Parameter Sets Validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{V2 - BestFit Parameter Sets Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE, warning=FALSE}
library(rfaR)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
```


## Purpose

Validate that the `qp3()` function and `lmom::cdfpe3()` correctly reproduce the volume-frequency curve (VFC) results from RMC-BestFit for John Martin Dam. This involves two tests:

1. **Credible Intervals** — Compute the 5th and 95th percentile flows across 10,000 BestFit LP3 parameter sets at each AEP and compare against the `jmd_vfc` dataset.
2. **Posterior Predictive** — For known discharges, compute the mean non-exceedance probability across all parameter sets and verify the resulting AEPs match `jmd_vfc`.

## Input Data

The `jmd_bf_parameter_sets` dataset contains 10,000 rows of LP3 parameters (mean, standard deviation, and skewness in log space) generated by RMC-BestFit via MCMC sampling. The `jmd_vfc` dataset contains the benchmark volume-frequency curve results including credible intervals and posterior predictive values.

```{r input-data}
knitr::kable(head(jmd_bf_parameter_sets, 5), digits = 4,
             col.names = c("Mean (log)", "SD (log)", "Skew (log)"),
             caption = "First 5 of 10,000 BestFit LP3 Parameter Sets")
```

```{r aep-table}
knitr::kable(data.frame(AEP = jmd_vfc$aep,
                        CI_95 = jmd_vfc$ci_95,
                        CI_5 = jmd_vfc$ci_5,
                        Posterior_Predictive = jmd_vfc$posterior_predictive),
             digits = 4,
             caption = "JMD Volume-Frequency Curve Benchmark Data")
```

---

## Test 1: Credible Intervals

For each AEP in `jmd_vfc`, compute the LP3 quantile (flow) from all 10,000 parameter sets using `qp3()`. Then take the 5th and 95th percentiles across parameter sets to form the credible interval. Compare against the benchmark values.

```{r test1-compute}
aeps <- jmd_vfc$aep

ci_matrix <- matrix(nrow = nrow(jmd_bf_parameter_sets), ncol = length(aeps))
for (i in 1:ncol(ci_matrix)) {
  for (j in 1:nrow(ci_matrix)) {
    ci_matrix[j, i] <- 10^qp3(1 - aeps[i],
                                jmd_bf_parameter_sets[j, 1],
                                jmd_bf_parameter_sets[j, 2],
                                jmd_bf_parameter_sets[j, 3])
  }
}

ci_5 <- numeric(ncol(ci_matrix))
ci_95 <- numeric(ncol(ci_matrix))
for (i in 1:ncol(ci_matrix)) {
  ci_5[i] <- as.numeric(quantile(ci_matrix[, i], 0.05))
  ci_95[i] <- as.numeric(quantile(ci_matrix[, i], 0.95))
}
```

```{r test1-table}
ci_results <- data.frame(
  AEP = aeps,
  Benchmark_CI95 = jmd_vfc$ci_95,
  Computed_CI95 = ci_95,
  Diff_CI95 = ci_95 - jmd_vfc$ci_95,
  Benchmark_CI5 = jmd_vfc$ci_5,
  Computed_CI5 = ci_5,
  Diff_CI5 = ci_5 - jmd_vfc$ci_5
)

knitr::kable(ci_results, digits = 4,
             col.names = c("AEP", "Benchmark 95%", "Computed 95%", "Diff 95%",
                           "Benchmark 5%", "Computed 5%", "Diff 5%"),
             caption = "Credible Interval Comparison (cfs)")
```

```{r test1-plot, fig.width=8, fig.height=6}
plot(aeps, jmd_vfc$ci_95, type = "l", col = "steelblue", lwd = 2,
     log = "xy", xlab = "Annual Exceedance Probability",
     ylab = "Flow (cfs)", main = "Credible Intervals: Benchmark vs. Computed",
     ylim = range(c(jmd_vfc$ci_5, jmd_vfc$ci_95)))
lines(aeps, jmd_vfc$ci_5, col = "steelblue", lwd = 2)
lines(aeps, ci_95, col = "coral", lwd = 2, lty = 2)
lines(aeps, ci_5, col = "coral", lwd = 2, lty = 2)
legend("topright", legend = c("Benchmark", "Computed"),
       col = c("steelblue", "coral"), lwd = 2, lty = c(1, 2))
```

### Acceptance Criterion

Maximum absolute difference must be less than 0.05 cfs for the 95th percentile and 0.01 cfs for the 5th percentile. Differences are attributable to interpolation methods between the Excel-based RMC-BestFit and direct R computation.

```{r test1-result}
max_diff_95 <- max(abs(ci_95 - jmd_vfc$ci_95))
max_diff_5 <- max(abs(ci_5 - jmd_vfc$ci_5))
pass_95 <- max_diff_95 < 0.05
pass_5 <- max_diff_5 < 0.01
```

| Metric | Value |
|--------|-------|
| Max Absolute Difference (95th CI) | `r round(max_diff_95, 6)` cfs |
| Tolerance (95th CI) | 0.05 cfs |
| **Result (95th CI)** | **`r ifelse(pass_95, "PASS", "FAIL")`** |
| Max Absolute Difference (5th CI) | `r round(max_diff_5, 6)` cfs |
| Tolerance (5th CI) | 0.01 cfs |
| **Result (5th CI)** | **`r ifelse(pass_5, "PASS", "FAIL")`** |

---

## Test 2: Posterior Predictive

For each discharge in `jmd_vfc$posterior_predictive`, compute the non-exceedance probability from all 10,000 parameter sets using `lmom::cdfpe3()`, then average to obtain the posterior predictive AEP. Compare against the benchmark AEPs.

```{r test2-compute}
discharges <- jmd_vfc$posterior_predictive
log_discharge <- log(discharges, base = 10)

posterior_matrix <- matrix(nrow = nrow(jmd_bf_parameter_sets), ncol = length(log_discharge))
for (i in 1:ncol(posterior_matrix)) {
  for (j in 1:nrow(posterior_matrix)) {
    nep <- lmom::cdfpe3(log_discharge[i],
                         c(jmd_bf_parameter_sets[j, 1],
                           jmd_bf_parameter_sets[j, 2],
                           jmd_bf_parameter_sets[j, 3]))
    posterior_matrix[j, i] <- nep
  }
}

post_pred <- numeric(ncol(posterior_matrix))
for (i in 1:ncol(posterior_matrix)) {
  post_pred[i] <- 1 - mean(posterior_matrix[, i])
}
```

```{r test2-table}
post_results <- data.frame(
  Discharge = discharges,
  Benchmark_AEP = aeps,
  Computed_AEP = post_pred,
  Difference = post_pred - aeps
)

knitr::kable(post_results, digits = 8,
             col.names = c("Discharge (cfs)", "Benchmark AEP", "Computed AEP", "Difference"),
             caption = "Posterior Predictive AEP Comparison")
```

### Acceptance Criterion

Maximum absolute difference between computed and benchmark AEPs must be less than 1e-4.

```{r test2-result}
max_diff_post <- max(abs(post_pred - aeps))
pass_post <- max_diff_post < 1e-4
```

| Metric | Value |
|--------|-------|
| Maximum Absolute Difference | `r format(max_diff_post, scientific = TRUE, digits = 4)` |
| Tolerance | 1e-4 |
| **Result** | **`r ifelse(pass_post, "PASS", "FAIL")`** |

---

## Summary

| Test | Description | Result |
|------|-------------|--------|
| 1a | 95th percentile credible interval | **`r ifelse(pass_95, "PASS", "FAIL")`** |
| 1b | 5th percentile credible interval | **`r ifelse(pass_5, "PASS", "FAIL")`** |
| 2 | Posterior predictive AEPs | **`r ifelse(pass_post, "PASS", "FAIL")`** |
