---
title: "Seasonality Sampling Validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Seasonality Sampling Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rfaR)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Purpose

Verify that sampling from the Jay McGraw Dam (JMD) seasonality distribution reproduces the input relative frequencies within expected Monte Carlo variability.

The seasonality module assigns each simulation realization to a calendar month based on the historical relative frequency of flood events. This test validates that `sample()` with the JMD seasonality probabilities produces a sampled distribution consistent with the input.

## Input Data

The input seasonality distribution is stored in `jmd_seasonality$relative_frequency`, a vector of 12 monthly probabilities that sum to 1.

```{r input-data}
seasonality_prob <- jmd_seasonality$relative_frequency

input_df <- data.frame(Month = month.abb,
                       Relative_Frequency = seasonality_prob)

knitr::kable(input_df, digits = 4, col.names = c("Month", "Relative Frequency"),
             caption = "JMD Seasonality Input Distribution")
```

## Test

Sample 100,000 months using the input probabilities, then compare the sampled relative frequencies to the input.

```{r sampling}
set.seed(42)
Nsims <- 100000

InitMonths <- sample(1:12, size = Nsims, replace = TRUE, prob = seasonality_prob)
sample_freq <- tabulate(InitMonths, nbins = 12) / Nsims
```

```{r results-table}
results_df <- data.frame(Month = month.abb,
                         Input = seasonality_prob,
                         Sampled = sample_freq,
                         Difference = sample_freq - seasonality_prob)

knitr::kable(results_df, digits = 4,
             col.names = c("Month", "Input Frequency", "Sampled Frequency", "Difference"),
             caption = "Input vs. Sampled Seasonality Frequencies (N = 100,000)")
```

```{r results-plot, fig.width=8, fig.height=5}
results_long <- results_df |>
  pivot_longer(cols = c(Input, Sampled),
               names_to  = "Type",
               values_to = "Frequency") |>
  mutate(Month = factor(Month, levels = month.abb),
         Type  = factor(Type, levels = c("Input", "Sampled")))

ggplot(results_long, aes(x = Month, y = Frequency, fill = Type)) +
  geom_col(position = "dodge", alpha = 0.8, color = "grey80") +
  scale_fill_manual(values = c("Input"   = "#3B4992FF",
                               "Sampled" = "#EE0000FF")) +
  labs(x = NULL,
       y = "Relative Frequency",
       fill = NULL) +
  scale_y_continuous(breaks = seq(0,0.5,0.1), minor_breaks = seq(0,0.5,0.01)) +
  theme_bw() +
  theme(legend.position  = "top",
        panel.grid.major.x = element_blank())
```

## Acceptance Criterion

Sampled relative frequencies must be within 1% relative tolerance of the input probabilities, consistent with `expect_equal(sample_freq, seasonality_prob, tolerance = 0.01)`.

```{r acceptance}
max_rel_diff <- max(abs((sample_freq - seasonality_prob) / seasonality_prob), na.rm = TRUE)
pass <- max_rel_diff < 0.01
```

| Metric | Value |
|--------|-------|
| Maximum Relative Difference | `r round(max_rel_diff, 6)` |
| Tolerance | 0.01 (1% relative) |
| **Result** | **`r ifelse(pass, "PASS", "FAIL")`** |
