---
title: "Hydrograph Scaling Validation"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Hydrograph Scaling Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.align = 'center',
  comment = "#>")

library(rfaR)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Purpose

Validate that `scale_hydrograph()` correctly scales inflow hydrographs and converts between timesteps. This test covers three cases:

1. **1-hour input, 1-hour routing** — Verify that the scale factor is applied correctly and output dimensions are preserved.
2. **15-minute input, 1-hour routing** — Verify that block averaging produces the correct timestep, row count, and ordinate values.
3. **Edge cases** — Verify behavior with a scale factor of 1 and non-negative flow output.

## Setup

```{r input-data}
hydrographs <- hydrograph_setup(jmd_hydro_apr1999,
                                jmd_hydro_jun1921,
                                jmd_hydro_jun1965,
                                jmd_hydro_jun1965_15min,
                                jmd_hydro_may1955,
                                jmd_hydro_pmf,
                                jmd_hydro_sdf,
                                critical_duration = 2,
                                routing_days = 10)

hydro_1hr   <- hydrographs[[1]]
hydro_15min <- hydrographs[[4]]

obs_vol    <- attr(hydro_1hr, "obs_vol")
obs_vol_15 <- attr(hydro_15min, "obs_vol")

sampled_vol_2x    <- obs_vol * 2
sampled_vol_15_2x <- obs_vol_15 * 2

```

```{r input-data-table, echo = FALSE}
input_summary <- data.frame(
  Hydrograph = c("April 1999 (1-hr)", "June 1965 (15-min)"),
  Timestep_hr = c(attr(hydro_1hr, "dt"), attr(hydro_15min, "dt")),
  Rows = c(nrow(hydro_1hr), nrow(hydro_15min)),
  Obs_Volume = c(obs_vol, obs_vol_15)
)

knitr::kable(input_summary, digits = 2,
             col.names = c("Hydrograph", "Timestep (hr)", "Rows", "Observed Volume"),
             caption = "Input Hydrograph Summary")
```
---

## Test 1: 1-Hour Input, 1-Hour Routing

Scale the April 1999 hydrograph by a factor of 2 with matching input and routing timesteps (1-hour).

```{r test1-compute}
scaled_1hr <- scale_hydrograph(
  hydrograph_shape = hydro_1hr[, c("hour", "inflow")],
  observed_volume  = obs_vol,
  sampled_volume   = sampled_vol_2x,
  routing_dt       = 1)
```

```{r test1-table, echo = FALSE}
test1_results <- data.frame(
  Check = c("Output class is data.frame",
            "Column names correct",
            "Time starts at 0",
            "Row count preserved",
            "Scale factor applied (all ordinates doubled)"),
  Result = c(
    is.data.frame(scaled_1hr),
    identical(names(scaled_1hr), c("time_hrs", "inflow_cfs")),
    scaled_1hr$time_hrs[1] == 0,
    nrow(scaled_1hr) == nrow(hydro_1hr),
    all.equal(scaled_1hr$inflow_cfs, hydro_1hr$inflow * 2, tolerance = 1e-10)
  )
)

knitr::kable(test1_results, caption = "1-Hour Input / 1-Hour Routing Checks")

```

```{r test1-plot, fig.width=8, fig.height=5, echo = FALSE}
hydro_long <- bind_rows(data.frame(time = hydro_1hr$hour,
                                   inflow = hydro_1hr$inflow,
                                   Type = "Original"),
                        data.frame(time = scaled_1hr$time_hrs,
                                   inflow = scaled_1hr$inflow_cfs,
                                   Type = "Scaled (2x)")) |>
  mutate(Type = factor(Type, levels = c("Original", "Scaled (2x)")))

ggplot(hydro_long, aes(x = time, y = inflow, color = Type, linetype = Type)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = c("Original"   = "#3B4992FF",
                                "Scaled (2x)" = "#EE0000FF")) +
  scale_linetype_manual(values = c("Original"    = "solid",
                                   "Scaled (2x)" = "solid")) +
  scale_y_continuous(label = scales::label_comma()) +
  scale_x_continuous(breaks = seq(0,240,24), minor_breaks = seq(0,240,12))+
  labs(x = "Time (hr)",
       y = "Inflow (cfs)",
       title = "1-Hour Hydrograph: Original vs. Scaled (2x)",
       color = NULL,
       linetype = NULL) +
  theme_bw() +
  theme(legend.position  = "top")
```


```{r test1-result, echo = FALSE}
is_df <- is.data.frame(scaled_1hr)
colnames_good <- identical(names(scaled_1hr), c("time_hrs", "inflow_cfs"))
zero_hour <- scaled_1hr$time_hrs[1] == 0
equal_rows <- nrow(scaled_1hr) == nrow(hydro_1hr)

pass_test1 <- all(
  is_df,
  colnames_good,
  zero_hour,
  equal_rows,
  isTRUE(all.equal(scaled_1hr$inflow_cfs, hydro_1hr$inflow * 2, tolerance = 1e-10)))
```

---

## Test 2: 15-Minute Input, 1-Hour Routing (Block Average)

Scale the June 1965 (15-min) hydrograph by a factor of 1 with block averaging to 1-hour routing timestep.

```{r test2-compute}
scaled_15min <- scale_hydrograph(
  hydrograph_shape = hydro_15min[, c("hour", "inflow")],
  observed_volume  = obs_vol_15,
  sampled_volume   = obs_vol_15,
  routing_dt       = 1)

expected_rows <- floor(nrow(hydro_15min) / 4)
dt_out <- scaled_15min$time_hrs[2] - scaled_15min$time_hrs[1]
expected_first <- mean(hydro_15min$inflow[1:4]) * 1
```

```{r test2-table, echo = FALSE}
test2_results <- data.frame(
  Check = c("Row count (15-min to 1-hr, factor of 4 reduction)",
            "Output timestep is 1 hour",
            "Time starts at 0",
            "First ordinate matches manual block average × scale"),
  Expected = c(as.character(expected_rows), "1", "0", round(expected_first, 4)),
  Actual = c(as.character(nrow(scaled_15min)),
             as.character(dt_out),
             as.character(scaled_15min$time_hrs[1]),
             round(scaled_15min$inflow_cfs[1], 4)),
  Pass = c(nrow(scaled_15min) == expected_rows,
           dt_out == 1,
           scaled_15min$time_hrs[1] == 0,
           isTRUE(all.equal(scaled_15min$inflow_cfs[1], expected_first, tolerance = 1e-10))))

knitr::kable(test2_results, caption = "15-Minute Input / 1-Hour Routing Checks")
```

```{r test2-plot, fig.width=8, fig.height=5, echo=FALSE, warning=FALSE}
hydro_15min_long <- bind_rows(data.frame(time   = hydro_15min$hour,
                                         inflow = hydro_15min$inflow,
                                         Type   = "Original (15-min)"),
                              data.frame(time   = scaled_15min$time_hrs,
                                         inflow = scaled_15min$inflow_cfs,
                                         Type   = "Scaled (1x, 1-hr)")) |>
  mutate(Type = factor(Type, levels = c("Original (15-min)", "Scaled (1x, 1-hr)")))

ggplot(hydro_15min_long) +
  geom_line(aes(x = time, y = inflow, 
                color = Type, 
                linetype = Type, 
                linewidth = Type)) +
  scale_color_manual(values = c("Original (15-min)" = "#3B4992FF",
                                "Scaled (1x, 1-hr)" = "#EE0000FF")) +
  scale_linetype_manual(values = c("Original (15-min)" = "solid",
                                   "Scaled (1x, 1-hr)" = "solid")) +
  scale_linewidth_manual(values = c("Original (15-min)" = 1.1,
                                   "Scaled (1x, 1-hr)" = 0.75)) +
  scale_y_continuous(label = scales::label_comma()) +
  scale_x_continuous(breaks = seq(0,240,24), minor_breaks = seq(0,240,12)) + 
  labs(x = "Time (hr)",
       y = "Inflow (cfs)",
       title = "15-Min Hydrograph: Original vs. Scaled (1x, Aggregated to 1-hr)",
       color = NULL,
       linetype = NULL,
       linewidth = NULL) +
  theme_bw() +
  theme(legend.position  = "top")
```

### Acceptance Criterion

```{r test2-result, echo = FALSE}
pass_test2 <- all(
  nrow(scaled_15min) == expected_rows,
  dt_out == 1,
  scaled_15min$time_hrs[1] == 0,
  isTRUE(all.equal(scaled_15min$inflow_cfs[1], expected_first, tolerance = 1e-10))
)
```


---

## Test 3: Edge Cases
**Unity scaling** `(sampled_volume = obs_vol)` verifies that when the sampled flood volume exactly matches the observed hydrograph volume, the scaling operation is a no-op — the output ordinates are identical to the input within floating point tolerance. This is an important baseline check: any scaling implementation that distorts the hydrograph shape when the scale factor is 1.0 has a fundamental error.

**Non-negative flows** verifies that the scaled hydrograph contains no negative inflow values. This can be a non-trivial failure mode when the scaling procedure involves interpolation, baseline correction, or any arithmetic that operates on near-zero recession limb ordinates. Negative inflows are physically meaningless and would corrupt the Modified Puls routing by implying water is being withdrawn from the reservoir by the inflow boundary condition.
```{r test3-compute}
# Scale factor = 1 (same volume)
scaled_unity <- scale_hydrograph(
  hydrograph_shape = hydro_1hr[, c("hour", "inflow")],
  observed_volume  = obs_vol,
  sampled_volume   = obs_vol,
  routing_dt       = 1
)

unity_match <- isTRUE(all.equal(scaled_unity$inflow_cfs, hydro_1hr$inflow, tolerance = 1e-10))

# Non-negative flows
scaled_2x <- scale_hydrograph(
  hydrograph_shape = hydro_1hr[, c("hour", "inflow")],
  observed_volume  = obs_vol,
  sampled_volume   = sampled_vol_2x,
  routing_dt       = 1
)

all_nonneg <- all(scaled_2x$inflow_cfs >= 0)
```

```{r test3-table, echo = FALSE}
test3_results <- data.frame(
  Check = c("Scale factor = 1 returns unchanged ordinates",
            "All flows non-negative for positive scale factor"),
  Result = c(unity_match, all_nonneg)
)

knitr::kable(test3_results, caption = "Edge Case Checks")
```

### Acceptance Criterion

```{r test3-result, echo = FALSE}
pass_test3 <- all(unity_match, all_nonneg)
```

---

## Summary

| Test | Description | Result |
|------|-------------|--------|
| 1 | 1-hour input, 1-hour routing — scaling | **`r ifelse(pass_test1, "PASS", "FAIL")`** |
| 2 | 15-min input, 1-hour routing — block average | **`r ifelse(pass_test2, "PASS", "FAIL")`** |
| 3 | Edge cases — unity scale, non-negative flows | **`r ifelse(pass_test3, "PASS", "FAIL")`** |
