---
title: "V7 - Modified-Puls Routing Validation"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{V7 - Modified-Puls Routing Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.align = 'center',
  comment = "#>")

library(rfaR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(scales)
```

## Purpose

Validate that `mod_puls_routing()` produces results consistent with HEC-HMS Modified Puls routing for a benchmark dataset at Cherry Cricket Dam. The comparison benchmark uses an identical reservoir model, inflow hydrograph, and initial condition routed independently in HEC-HMS.

---

## Modified Puls Routing

The Modified Puls routing method, also known as storage routing or level-pool routing, is based upon a finite difference approximation of the continuity equation, coupled with an empirical representation of the momentum equation (Chow, 1964; Henderson, 1966; HEC, 2025). The full derivation of the Mod-Puls routing equation from the continuity equation can be found in the rfaR Reference Manual.

Mod-Puls routing uses the known inflow, outflow, and storage at time *t-1* and the known inflow at time *t* to estimate the the storage indicator at time *t*. The storage indicator uses the stage-storage-discharge relationship (`your_res_model`) to determine the outflow at time *t*.

$$SI_t = \frac{I_{t-1} + I_t}{2} + \left(\frac{S_{t-1}}{\Delta t} - \frac{O_{t-1}}{2}\right)$$

---

## Test

```{r cc-compute}
cc_routing <- mod_puls_routing(resmodel_df = cc_resmodel,
                               inflow_df = cc_inflowhydro,
                               initial_elev = cc_init_elev,
                               full_results = TRUE)

cc_diff_elev <- cc_routing$elevation_ft - cc_hms_results$elevation_ft
cc_diff_outflow <- cc_routing$outflow_cfs - cc_hms_results$outflow_cfs

cc_max_diff_elev <- max(abs(cc_diff_elev))
cc_max_diff_outflow <- max(abs(cc_diff_outflow))
```

```{r cc-table, echo = FALSE}
cc_summary <- data.frame(
  Metric = c("Peak Elevation (ft)", "Peak Outflow (cfs)",
             "Max Abs Diff Elevation (ft)", "Max Abs Diff Outflow (cfs)"),
  rfaR = c(max(cc_routing$elevation_ft), max(cc_routing$outflow_cfs),
           cc_max_diff_elev, cc_max_diff_outflow),
  HMS = c(max(cc_hms_results$elevation_ft), max(cc_hms_results$outflow_cfs),
          NA, NA)
)

knitr::kable(cc_summary, digits = 4,
             col.names = c("Metric", "rfaR", "HEC-HMS"),
             caption = "Cherry Creek Dam Routing Comparison")
```

```{r cc-plot, fig.width=10, fig.height=8, echo = FALSE}
hydro_compare_long <- bind_rows(
  data.frame(time   = cc_routing$time_hr,
             flow   = cc_routing$inflow_cfs,
             Series = "Inflow"),
  data.frame(time   = cc_hms_results$time_hr,
             flow   = cc_hms_results$outflow_cfs,
             Series = "HMS Outflow"),
  data.frame(time   = cc_routing$time_hr,
             flow   = cc_routing$outflow_cfs,
             Series = "rfaR Outflow")) |>
  mutate(Series = factor(Series, levels = c("Inflow", "HMS Outflow", "rfaR Outflow")))

# Difference data
diff_df <- data.frame(time = cc_routing$time_hr,
                      diff = cc_diff_outflow)

# Top panel
p1 <- ggplot(hydro_compare_long, aes(x = time, y = flow,
                                     color = Series, linetype = Series,
                                     linewidth = Series)) +
  geom_line() +
  scale_color_manual(values = c("Inflow" = "gray70",
                                "rfaR Outflow" = "#EE0000FF",
                                "HMS Outflow" = "#3B4992FF")) +
  scale_linetype_manual(values = c("Inflow"  = "solid",
                                   "rfaR Outflow" = "dashed",
                                   "HMS Outflow" = "solid")) +
  scale_linewidth_manual(values = c("Inflow" = 0.6,
                                    "rfaR Outflow" = 1.0,
                                    "HMS Outflow" = 1.0)) +
  scale_x_continuous(breaks = seq(0,960,24)) +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(x = "Time (hr)",
       y = "Flow (cfs)",
       title = "Cherry Cricket Dam \u2014 Inflow and Outflow Comparison",
       color = NULL,
       linetype = NULL,
       linewidth = NULL) +
  theme_bw() +
  theme(legend.position  = "inside",
        legend.position.inside = c(0.8,0.8)) + 
  coord_cartesian(ylim = c(0,2500))

# Bottom panel
p2 <- ggplot(diff_df) +
  geom_histogram(aes(x = diff),
                 binwidth = 5e-6,
                 color = "grey70",
                 fill = "#008280FF",
                 alpha = 0.75) + 
  scale_x_continuous(breaks = scales::breaks_pretty(n = 10))+
  geom_vline(xintercept = 0, color = "gray50", linetype = "dashed") +
  labs(x = "Cherry Cricket Dam — Outflow Difference (rfaR − HMS)",
       y = "Frequency",
       title = "Distribution of Calculation Differences") +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

p1 / p2
```

### Acceptance Criterion

Maximum absolute difference must be less than 0.1 ft for elevation and 1.0 cfs for outflow.

```{r cc-result, echo = FALSE}
pass_cc_elev <- cc_max_diff_elev < 0.1
pass_cc_outflow <- cc_max_diff_outflow < 1.0
```

| Metric | Value | Tolerance |
|--------|-------|-----------|
| Max Abs Diff Elevation | `r round(cc_max_diff_elev, 4)` ft | 0.1 ft |
| Max Abs Diff Outflow | `r round(cc_max_diff_outflow, 4)` cfs | 1.0 cfs |
| **Result (Elevation)** | **`r ifelse(pass_cc_elev, "PASS", "FAIL")`** | |
| **Result (Outflow)** | **`r ifelse(pass_cc_outflow, "PASS", "FAIL")`** | |

---

## Summary

| Test | Dam | Variable | Result |
|------|-----|----------|--------|
| 1a | Cherry Cricket | Elevation | **`r ifelse(pass_cc_elev, "PASS", "FAIL")`** |
| 1b | Cherry Cricket | Outflow | **`r ifelse(pass_cc_outflow, "PASS", "FAIL")`** |

## References

U.S. Army Corps of Engineers, Hydrologic Engineering Center. (n.d.). Modified Puls Model. HEC-HMS Technical Reference Manual. Retrieved December 18, 2025, from https://www.hec.usace.army.mil/confluence/hmsdocs/hmstrm/channel-flow/modified-puls-model
