---
title: "Validation of the Modified-Puls Routing Function"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation of the Modified-Puls Routing Function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rfaR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(scales)
```

## Purpose

Validate that `mod_puls_routing()` produces results consistent with HEC-HMS Modified Puls routing for a benchmark dataset at Cherry Cricket Dam. The comparison benchmark uses an identical reservoir model, inflow hydrograph, and initial condition routed independently in HEC-HMS.

---

## Modified Puls Routing

The Modified Puls routing method, also known as storage routing or level-pool routing, is based upon a finite difference approximation of the continuity equation, coupled with an empirical representation of the momentum equation (Chow, 1964; Henderson, 1966; HEC, 2025).

The continuity equation for a control volume is:

$$\frac{\partial Q}{\partial x} + \frac{\partial A}{\partial t} = 0$$

where *Q* is discharge, *A* is cross-sectional area, *x* is distance, and *t* is time. For reservoir routing, this is expressed using average inflow and outflow over a time step:

$$\bar{I_t} - \bar{O_t} = \frac{\Delta S_t}{\Delta t}$$

where $\bar{I_t}$ is the average inflow, $\bar{O_t}$ is the average outflow, and $\Delta S_t$ is the change in storage over the time interval $\Delta t$.

Rearranging and expressing in finite difference form yields the Modified Puls routing equation:

$$\left(\frac{S_t}{\Delta t} + \frac{O_t}{2}\right) = \frac{I_{t-1} + I_t}{2} + \left(\frac{S_{t-1}}{\Delta t} - \frac{O_{t-1}}{2}\right)$$

The method introduces a "storage indication" term:

$$SI_t = \frac{S_t}{\Delta t} + \frac{O_t}{2}$$

Substituting this into the routing equation:

$$SI_t = \frac{I_{t-1} + I_t}{2} + \left(\frac{S_{t-1}}{\Delta t} - \frac{O_{t-1}}{2}\right)$$

The left side represents the storage indication at time *t*, which can be computed from known values on the right side. This value is then used with the storage-outflow relationship to determine the outflow at time *t*.

---

## Test

```{r cc-compute}
cc_routing <- mod_puls_routing(resmodel_df = cc_resmodel,
                               inflow_df = cc_inflowhydro,
                               initial_elev = cc_init_elev,
                               full_results = TRUE)

cc_diff_elev <- cc_routing$elevation_ft - cc_hms_results$elevation_ft
cc_diff_outflow <- cc_routing$outflow_cfs - cc_hms_results$outflow_cfs

cc_max_diff_elev <- max(abs(cc_diff_elev))
cc_max_diff_outflow <- max(abs(cc_diff_outflow))
```

```{r cc-table}
cc_summary <- data.frame(
  Metric = c("Peak Elevation (ft)", "Peak Outflow (cfs)",
             "Max Abs Diff Elevation (ft)", "Max Abs Diff Outflow (cfs)"),
  rfaR = c(max(cc_routing$elevation_ft), max(cc_routing$outflow_cfs),
           cc_max_diff_elev, cc_max_diff_outflow),
  HMS = c(max(cc_hms_results$elevation_ft), max(cc_hms_results$outflow_cfs),
          NA, NA)
)

knitr::kable(cc_summary, digits = 4,
             col.names = c("Metric", "rfaR", "HEC-HMS"),
             caption = "Cherry Creek Dam Routing Comparison")
```

```{r cc-plot, fig.width=10, fig.height=8}
hydro_compare_long <- bind_rows(
  data.frame(time   = cc_routing$time_hr,
             flow   = cc_routing$inflow_cfs,
             Series = "Inflow"),
  data.frame(time   = cc_hms_results$time_hr,
             flow   = cc_hms_results$outflow_cfs,
             Series = "HMS Outflow"),
  data.frame(time   = cc_routing$time_hr,
             flow   = cc_routing$outflow_cfs,
             Series = "rfaR Outflow")) |>
  mutate(Series = factor(Series, levels = c("Inflow", "HMS Outflow", "rfaR Outflow")))

# Difference data
diff_df <- data.frame(time = cc_routing$time_hr,
                      diff = cc_diff_outflow)

# Top panel
p1 <- ggplot(hydro_compare_long, aes(x = time, y = flow,
                                     color = Series, linetype = Series,
                                     linewidth = Series)) +
  geom_line() +
  scale_color_manual(values = c("Inflow" = "gray70",
                                "rfaR Outflow" = "#EE0000FF",
                                "HMS Outflow" = "#3B4992FF")) +
  scale_linetype_manual(values = c("Inflow"  = "solid",
                                   "rfaR Outflow" = "dashed",
                                   "HMS Outflow" = "solid")) +
  scale_linewidth_manual(values = c("Inflow" = 0.6,
                                    "rfaR Outflow" = 1.0,
                                    "HMS Outflow" = 1.0)) +
  scale_x_continuous(breaks = seq(0,960,24)) +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(x = "Time (hr)",
       y = "Flow (cfs)",
       title = "Cherry Cricket Dam \u2014 Inflow and Outflow Comparison",
       color = NULL,
       linetype = NULL,
       linewidth = NULL) +
  theme_bw() +
  theme(legend.position  = "top") + 
  coord_cartesian(ylim = c(0,2500))

# Bottom panel
p2 <- ggplot(diff_df, aes(x = time, y = diff)) +
  geom_line(color = "#3B4992FF", linewidth = 0.6) +
  geom_hline(yintercept = 0, color = "gray50", linetype = "dashed") +
  labs(x     = "Time (hr)",
       y     = "Difference (cfs)",
       title = "Cherry Cricket Dam \u2014 Outflow Difference (rfaR \u2212 HMS)") +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

p1 / p2
```

### Acceptance Criterion

Maximum absolute difference must be less than 0.1 ft for elevation and 1.0 cfs for outflow.

```{r cc-result}
pass_cc_elev <- cc_max_diff_elev < 0.1
pass_cc_outflow <- cc_max_diff_outflow < 1.0
```

| Metric | Value | Tolerance |
|--------|-------|-----------|
| Max Abs Diff Elevation | `r round(cc_max_diff_elev, 4)` ft | 0.1 ft |
| Max Abs Diff Outflow | `r round(cc_max_diff_outflow, 4)` cfs | 1.0 cfs |
| **Result (Elevation)** | **`r ifelse(pass_cc_elev, "PASS", "FAIL")`** | |
| **Result (Outflow)** | **`r ifelse(pass_cc_outflow, "PASS", "FAIL")`** | |

---

## Summary

| Test | Dam | Variable | Result |
|------|-----|----------|--------|
| 1a | Cherry Cricket | Elevation | **`r ifelse(pass_cc_elev, "PASS", "FAIL")`** |
| 1b | Cherry Cricket | Outflow | **`r ifelse(pass_cc_outflow, "PASS", "FAIL")`** |

## References
Chow, V. T. (1964). Handbook of Applied Hydrology. McGraw-Hill Book Company, New York.

Henderson, F. M. (1966). Open Channel Flow. Macmillan Publishing Co., Inc., New York.

Nash, J. E., & Sutcliffe, J. V. (1970). River flow forecasting through conceptual models part I — A discussion of principles. Journal of Hydrology, 10(3), 282-290. https://doi.org/10.1016/0022-1694(70)90255-6

U.S. Army Corps of Engineers, Hydrologic Engineering Center. (n.d.). Modified Puls Model. HEC-HMS Technical Reference Manual. Retrieved December 18, 2025, from https://www.hec.usace.army.mil/confluence/hmsdocs/hmstrm/channel-flow/modified-puls-model
