---
title: "Hydrograph Shape Sampling Validation"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Hydrograph Shape Sampling Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.align = 'center',
  comment = "#>")

library(rfaR)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
```

## Purpose

Validate that the hydrograph shape sampling module correctly applies user-specified weights when selecting hydrograph shapes during simulation. This test validates that `hydrograph_setup()` correctly normalizes weights to probabilities and that `sample()` using those probabilities reproduces the expected distribution.

## Input Data

Seven example hydrographs are loaded via `hydrograph_setup()` with the following user-specified weights:

```{r input-data}
hydrograph_names <- c("April 1999", "June 1921", "June 1965 (1hr)",
                      "June 1965 (15min)", "May 1955", "PMF", "SDF")
weights <- c(0.5, 1, 1, 2, 2, 2, 4)
expected_probs <- weights / sum(weights)

```


```{r input-data-table, echo = FALSE}
input_df <- data.frame(
  Hydrograph = hydrograph_names,
  Weight = weights,
  Expected_Probability = expected_probs,
  exp_prob_equation = paste0(weights,"/", sum(weights)))

knitr::kable(input_df, digits = 4,
             col.names = c("Hydrograph", "Weight", "Expected Prob.", "Expected Prob. Fraction"),
             caption = "Hydrograph Sampling Weights and Normalized Probabilities")
```

## Test

Set up hydrographs with weights, then sample 1,000,000 times and compare the sampled frequencies to the expected probabilities.

```{r sampling}
set.seed(42)

hydrographs <- hydrograph_setup(jmd_hydro_apr1999,
                                jmd_hydro_jun1921,
                                jmd_hydro_jun1965,
                                jmd_hydro_jun1965_15min,
                                jmd_hydro_may1955,
                                jmd_hydro_pmf,
                                jmd_hydro_sdf,
                                critical_duration = 2,
                                routing_days = 10,
                                weights = weights)

Nsims <- 1000000
hydro_probs <- attr(hydrographs, "probs")
hydroSamps <- sample(1:length(hydrographs), size = Nsims, replace = TRUE,
                     prob = hydro_probs)
sample_freq <- tabulate(hydroSamps, nbins = length(hydro_probs)) / Nsims
```

```{r results-table, echo = FALSE}
results_df <- data.frame(
  Hydrograph = hydrograph_names,
  Weight = weights,
  Expected = expected_probs,
  Sampled = sample_freq,
  Difference = sample_freq - expected_probs
)

knitr::kable(results_df, digits = 4,
             col.names = c("Hydrograph", "Weight", "Expected Prob.", "Sampled Freq.", "Difference"),
             caption = "Expected vs. Sampled Hydrograph Frequencies (N = 1,000,000)")
```

```{r results-plot, fig.width=8, fig.height=5, echo = FALSE}
results_long <- results_df |>
  pivot_longer(cols = c(Expected, Sampled),
               names_to = "Type",
               values_to = "Probability") |>
  mutate(Hydrograph = factor(Hydrograph, levels = hydrograph_names),
         Type       = factor(Type, levels = c("Expected", "Sampled")))

ggplot(results_long, aes(x = Hydrograph, y = Probability, fill = Type)) +
  geom_col(position = "dodge", alpha = 0.8, color = "grey80") +
  scale_fill_manual(values = c("Expected" = "#3B4992FF",
                               "Sampled"  = "#EE0000FF")) +
  labs(x = NULL,
       y = "Probability",
       title = "Hydrograph Sampling: Expected vs. Sampled Frequencies",
       fill = NULL) +
  scale_y_continuous(breaks = seq(0,0.5,0.1), minor_breaks = seq(0,0.5,0.025)) +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major.x = element_blank())
```

```{r nsim decay, echo = FALSE, warning = FALSE}
set.seed(42)
power_funct <- function(x) {
  as.integer(outer(c(1,2.5,5,7.5), 10^(1:x)))
}

nsim_vect <- power_funct(8)

nsim_df <- do.call(rbind, lapply(nsim_vect, \(NSims) {
  
  hydroSamps <- sample(1:length(hydrographs), size = NSims, replace = TRUE,
                     prob = hydro_probs)
  sample_freq <- tabulate(hydroSamps, nbins = length(hydro_probs)) / NSims
  
  data.frame(
    Nsims   = NSims,
    Input   = expected_probs,
    Sampled = sample_freq,
    Diff    = sample_freq - expected_probs,
    PctDiff = (sample_freq - expected_probs) / expected_probs * 100
  )
}))

decay_curve <- nsim_df |> 
  group_by(Nsims) |> 
  slice_max(abs(Diff),n = 1)

ggplot(decay_curve) + 
  geom_line(aes(x = Nsims, y = abs(Diff)), color =  "#008280FF", linewidth = 0.8) + 
  scale_x_log10(breaks = scales::breaks_log(n = 10,base = 10),
                minor_breaks = minor_breaks_log(detail = 1),
                labels = scales::label_scientific(digits = 1)) + 
  theme_bw() + 
  labs(y = "Abs. Difference (sample - expected)",
       x = "Number of Simulations (Nsims)",
       title = "Absolute Difference of Sampled and Expected Relative Frequency",
       subtitle = "Convergence to input frequencies occurs with increasing number of simulations")
```
## Acceptance Criterion

Sampled frequencies must be within 1% relative tolerance of the expected probabilities, consistent with `expect_equal(sample_freq, hydro_probs, tolerance = 0.01)`.

```{r acceptance, echo = FALSE}
max_rel_diff <- max(abs((sample_freq - expected_probs) / expected_probs), na.rm = TRUE)
pass <- max_rel_diff < 0.01
```

| Metric | Value |
|--------|-------|
| Maximum Relative Difference | `r round(max_rel_diff, 6)` |
| Tolerance | 0.01 (1% relative) |
| **Result** | **`r ifelse(pass, "PASS", "FAIL")`** |
