---
title: "Hydrograph Shape Sampling Validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hydrograph Shape Sampling Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rfaR)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Purpose

Verify that the hydrograph shape sampling module correctly applies user-specified weights when selecting hydrograph shapes during simulation. This test validates that `hydrograph_setup()` correctly normalizes weights to probabilities and that `sample()` using those probabilities reproduces the expected distribution.

## Input Data

Seven JMD hydrographs are loaded via `hydrograph_setup()` with the following user-specified weights:

```{r input-data}
hydrograph_names <- c("April 1999", "June 1921", "June 1965 (1hr)",
                      "June 1965 (15min)", "May 1955", "PMF", "SDF")
weights <- c(0.5, 1, 1, 2, 2, 2, 4)
expected_probs <- weights / sum(weights)

input_df <- data.frame(
  Hydrograph = hydrograph_names,
  Weight = weights,
  Expected_Probability = expected_probs
)

knitr::kable(input_df, digits = 4,
             col.names = c("Hydrograph", "Weight", "Expected Probability"),
             caption = "Hydrograph Sampling Weights and Normalized Probabilities")
```

## Test

Set up hydrographs with weights, then sample 100,000 times and compare the sampled frequencies to the expected probabilities.

```{r sampling}
set.seed(42)

hydrographs <- hydrograph_setup(jmd_hydro_apr1999,
                                jmd_hydro_jun1921,
                                jmd_hydro_jun1965,
                                jmd_hydro_jun1965_15min,
                                jmd_hydro_may1955,
                                jmd_hydro_pmf,
                                jmd_hydro_sdf,
                                critical_duration = 2,
                                routing_days = 10,
                                weights = weights)

Nsims <- 1000000
hydro_probs <- attr(hydrographs, "probs")
hydroSamps <- sample(1:length(hydrographs), size = Nsims, replace = TRUE,
                     prob = hydro_probs)
sample_freq <- tabulate(hydroSamps, nbins = length(hydro_probs)) / Nsims
```

```{r results-table}
results_df <- data.frame(
  Hydrograph = hydrograph_names,
  Weight = weights,
  Expected = expected_probs,
  Sampled = sample_freq,
  Difference = sample_freq - expected_probs
)

knitr::kable(results_df, digits = 4,
             col.names = c("Hydrograph", "Weight", "Expected Prob.", "Sampled Freq.", "Difference"),
             caption = "Expected vs. Sampled Hydrograph Frequencies (N = 100,000)")
```

```{r results-plot, fig.width=8, fig.height=5}
results_long <- results_df |>
  pivot_longer(cols = c(Expected, Sampled),
               names_to = "Type",
               values_to = "Probability") |>
  mutate(Hydrograph = factor(Hydrograph, levels = hydrograph_names),
         Type       = factor(Type, levels = c("Expected", "Sampled")))

ggplot(results_long, aes(x = Hydrograph, y = Probability, fill = Type)) +
  geom_col(position = "dodge", alpha = 0.8, color = "grey80") +
  scale_fill_manual(values = c("Expected" = "#3B4992FF",
                               "Sampled"  = "#EE0000FF")) +
  labs(x = NULL,
       y = "Probability",
       title = "Hydrograph Sampling: Expected vs. Sampled Frequencies",
       fill = NULL) +
  scale_y_continuous(breaks = seq(0,0.5,0.1), minor_breaks = seq(0,0.5,0.025)) +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major.x = element_blank())
```

## Acceptance Criterion

Sampled frequencies must be within 1% relative tolerance of the expected probabilities, consistent with `expect_equal(sample_freq, hydro_probs, tolerance = 0.01)`.

```{r acceptance}
max_rel_diff <- max(abs((sample_freq - expected_probs) / expected_probs), na.rm = TRUE)
pass <- max_rel_diff < 0.01
```

| Metric | Value |
|--------|-------|
| Maximum Relative Difference | `r round(max_rel_diff, 6)` |
| Tolerance | 0.01 (1% relative) |
| **Result** | **`r ifelse(pass, "PASS", "FAIL")`** |
