---
title: "Validation: Seasonality Sampling — John Martin Dam"
author: "rfaR Package Validation"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(rfaR)
library(dplyr)
```

## Purpose

Verify that sampling from the John Martin Dam (JMD) seasonality distribution reproduces the input relative frequencies within expected Monte Carlo variability.

The seasonality module assigns each simulation realization to a calendar month based on the historical relative frequency of flood events. This test validates that `sample()` with the JMD seasonality probabilities produces a sampled distribution consistent with the input.

## Input Data

The input seasonality distribution is stored in `jmd_seasonality$relative_frequency`, a vector of 12 monthly probabilities that sum to 1.

```{r input-data}
seasonality_prob <- jmd_seasonality$relative_frequency

input_df <- data.frame(
  Month = month.abb,
  Relative_Frequency = seasonality_prob
)

knitr::kable(input_df, digits = 4, col.names = c("Month", "Relative Frequency"),
             caption = "JMD Seasonality Input Distribution")
```

## Test

Sample 100,000 months using the input probabilities, then compare the sampled relative frequencies to the input.

```{r sampling}
set.seed(42)
Nsims <- 100000

InitMonths <- sample(1:12, size = Nsims, replace = TRUE, prob = seasonality_prob)
sample_freq <- tabulate(InitMonths, nbins = 12) / Nsims
```

```{r results-table}
results_df <- data.frame(
  Month = month.abb,
  Input = seasonality_prob,
  Sampled = sample_freq,
  Difference = sample_freq - seasonality_prob
)

knitr::kable(results_df, digits = 4,
             col.names = c("Month", "Input Frequency", "Sampled Frequency", "Difference"),
             caption = "Input vs. Sampled Seasonality Frequencies (N = 100,000)")
```

```{r results-plot, fig.width=8, fig.height=5}
barplot(rbind(results_df$Input, results_df$Sampled),
        beside = TRUE, names.arg = results_df$Month,
        col = c("steelblue", "coral"),
        ylab = "Relative Frequency", main = "Seasonality: Input vs. Sampled Frequencies",
        legend.text = c("Input", "Sampled"),
        args.legend = list(x = "topright"))
```

## Acceptance Criterion

Sampled relative frequencies must be within 1% relative tolerance of the input probabilities, consistent with `expect_equal(sample_freq, seasonality_prob, tolerance = 0.01)`.

```{r acceptance}
max_rel_diff <- max(abs((sample_freq - seasonality_prob) / seasonality_prob), na.rm = TRUE)
pass <- max_rel_diff < 0.01
```

| Metric | Value |
|--------|-------|
| Maximum Relative Difference | `r round(max_rel_diff, 6)` |
| Tolerance | 0.01 (1% relative) |
| **Result** | **`r ifelse(pass, "PASS", "FAIL")`** |
