---
title: "Validation: Hydrograph Shape Sampling — John Martin Dam"
author: "rfaR Package Validation"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(rfaR)
library(dplyr)
```

## Purpose

Verify that the hydrograph shape sampling module correctly applies user-specified weights when selecting hydrograph shapes during simulation. This test validates that `hydrograph_setup()` correctly normalizes weights to probabilities and that `sample()` using those probabilities reproduces the expected distribution.

## Input Data

Seven JMD hydrographs are loaded via `hydrograph_setup()` with the following user-specified weights:

```{r input-data}
hydrograph_names <- c("April 1999", "June 1921", "June 1965 (1hr)",
                      "June 1965 (15min)", "May 1955", "PMF", "SDF")
weights <- c(0.5, 1, 1, 2, 2, 2, 4)
expected_probs <- weights / sum(weights)

input_df <- data.frame(
  Hydrograph = hydrograph_names,
  Weight = weights,
  Expected_Probability = expected_probs
)

knitr::kable(input_df, digits = 4,
             col.names = c("Hydrograph", "Weight", "Expected Probability"),
             caption = "Hydrograph Sampling Weights and Normalized Probabilities")
```

## Test

Set up hydrographs with weights, then sample 100,000 times and compare the sampled frequencies to the expected probabilities.

```{r sampling}
set.seed(42)

hydrographs <- hydrograph_setup(jmd_hydro_apr1999,
                                jmd_hydro_jun1921,
                                jmd_hydro_jun1965,
                                jmd_hydro_jun1965_15min,
                                jmd_hydro_may1955,
                                jmd_hydro_pmf,
                                jmd_hydro_sdf,
                                critical_duration = 2,
                                routing_days = 10,
                                weights = weights)

Nsims <- 100000
hydro_probs <- attr(hydrographs, "probs")
hydroSamps <- sample(1:length(hydrographs), size = Nsims, replace = TRUE,
                     prob = hydro_probs)
sample_freq <- tabulate(hydroSamps, nbins = length(hydro_probs)) / Nsims
```

```{r results-table}
results_df <- data.frame(
  Hydrograph = hydrograph_names,
  Weight = weights,
  Expected = expected_probs,
  Sampled = sample_freq,
  Difference = sample_freq - expected_probs
)

knitr::kable(results_df, digits = 4,
             col.names = c("Hydrograph", "Weight", "Expected Prob.", "Sampled Freq.", "Difference"),
             caption = "Expected vs. Sampled Hydrograph Frequencies (N = 100,000)")
```

```{r results-plot, fig.width=8, fig.height=5}
barplot(rbind(results_df$Expected, results_df$Sampled),
        beside = TRUE, names.arg = hydrograph_names,
        col = c("steelblue", "coral"), las = 2, cex.names = 0.8,
        ylab = "Probability", main = "Hydrograph Sampling: Expected vs. Sampled Frequencies",
        legend.text = c("Expected", "Sampled"),
        args.legend = list(x = "topright"))
```

## Acceptance Criterion

Sampled frequencies must be within 1% relative tolerance of the expected probabilities, consistent with `expect_equal(sample_freq, hydro_probs, tolerance = 0.01)`.

```{r acceptance}
max_rel_diff <- max(abs((sample_freq - expected_probs) / expected_probs), na.rm = TRUE)
pass <- max_rel_diff < 0.01
```

| Metric | Value |
|--------|-------|
| Maximum Relative Difference | `r round(max_rel_diff, 6)` |
| Tolerance | 0.01 (1% relative) |
| **Result** | **`r ifelse(pass, "PASS", "FAIL")`** |
